---
output:
  pdf_document: default
  html_document: default
---
## Section 4: Population Scale Analysis

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes
  
```{r}
expr <- read.table("class12_file.txt")
  head(expr)
```
```{r}
nrow(expr)
```


```{r}
# Sample size for each genotype
table(expr$geno)
```

```{r}
# Median expression levels for A/A
  median(expr$exp[expr$geno == "A/A"])
```

```{r}
# Median expression levels for A/G
  median(expr$exp[expr$geno == "A/G"])
```

```{r}
# Median expression levels for G/G
  median(expr$exp[expr$geno == "G/G"])
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2) 
ggplot(expr) + aes(geno, exp, fill = geno) + geom_boxplot(notch = TRUE)
```

Expression increases with the A allele or decreases with the G allele but the overall effect is not significatn because the expression levels are all overlapping so there is no significant difference




