---
title: "Class05: Data Vis with ggplot"
author: "James Garza (PID: A16300772)"
format: pdf
---

# Graphics systems in R

There are many graphics systems in R for making plots and figures.

We have alreasy played a little with **"base R"** graphics and the `plot()` function

Today we will start learning about a popular graphics package called `ggplot2()`.

This is an add on package - i.e. we need to install it.  I install it (like I install any package) with the `install.package()` function.

```{r}
# install.packages("ggplot2")
# library(ggplot2)
```


```{r}
plot(cars)
```

Before I can use the funstions from a package I have to load up the package from my "library". We use the `library(ggplot2)` command to load it up.

```{r}
library(ggplot2)
ggplot(cars)
```

Every ggplot is made up of at least 3 things: 
- data (the numbers etc. that will go into your plot)
- aes (how the columns of ata map to the plot aesthetics)
- geoms (how the plot actually looks, points, bars, lines etc.)

"Command" "Option" "i" on Mac to do shortcut to insert coding area


```{r}
ggplot(cars) + aes(x=speed, y=dist) + geom_point()
```

For simple plots ggplot is mroe verbose - it takes more code - than base R plot.
Add some more layers to our ggplot:

```{r}
ggplot(cars) + aes(x=speed, y=dist) + geom_point() + geom_smooth(method="lm",level=0) + labs(title="Stopping distance of old cars", subtitle= "A silly example plot") + theme_bw()
```

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
nrow(genes)
colnames(genes)
ncol(genes)
```
```{r}
round( table(genes$State)/nrow(genes) * 100, 2 )
```
```{r}
p <- ggplot(genes) + 
    aes(x=Condition1, y=Condition2, col=State) + geom_point()
p
```
```{r}
p + scale_colour_manual( values=c("blue","gray","red") )
```
```{r}
p + labs(title="Gene Expression Changes Upon Drug Treatment", x="Control (no drug)", y= "Drug Treatment")
```
```{r}
p + labs(title="Gene Expression Changes Upon Drug Treatment", x="Control (no drug)", y= "Drug Treatment") + scale_colour_manual(values=c("blue","black","red"))

```
```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```
```{r}
library(dplyr)

gapminder_2007 <- gapminder %>% filter(year==2007)
```
```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp) +
  geom_point()
```
```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp) +
  geom_point(alpha=0.5)
```
```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp, color=continent, size=pop) +
  geom_point(alpha=0.5)
```
```{r}
ggplot(gapminder_2007) + 
  aes(x = gdpPercap, y = lifeExp, color = pop) +
  geom_point(alpha=0.8)
```
```{r}
ggplot(gapminder_2007) + 
  aes(x = gdpPercap, y = lifeExp, size = pop) +
  geom_point(alpha=0.5)
```
```{r}
ggplot(gapminder_2007) + 
  geom_point(aes(x = gdpPercap, y = lifeExp,
                 size = pop), alpha=0.5) + 
  scale_size_area(max_size = 10)
```
```{r}
gapminder_1957 <- gapminder %>% filter(year==1957)
```
```{r}
p2 <- ggplot(gapminder_1957) +
  aes(x=gdpPercap, y=lifeExp) +
  geom_point()
p2
```
```{r}
ggplot(gapminder_1957) + 
  geom_point(aes(x = gdpPercap, y = lifeExp,
                 size = pop, color=continent), alpha=0.7) + 
  scale_size_area(max_size = 15)
```
```{r}
gapminder_1957 <- gapminder %>% filter(year==1957 | year==2007)

ggplot(gapminder_1957) + 
  geom_point(aes(x = gdpPercap, y = lifeExp, color=continent,
                 size = pop), alpha=0.7) + 
  scale_size_area(max_size = 10) +
  facet_wrap(~year)
```










